---
title: "seqNdisplayR"
author: "Søren Lykke-Andersen, Jerome Rouviere, Manfred Schmid, Maria Gockert, Torben Heick Jensen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{seqNdisplayR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # Default all chunks to non-evaluating so vignette builds don't fetch
  # remote bigwigs. To run the vignette interactively, either flip this
  # to TRUE or run individual chunks by hand.
  eval = FALSE
)
```

## Introduction

`seqNdisplayR` produces customisable, reproducible genome-coverage plots
for comparative sequencing visualisation.  There are two usage paths:

- **Shiny app (primary).**  A point-and-click interface for loading an
  Excel sample sheet, choosing a locus, tweaking layout, and drawing a
  publication-ready plot.  Most users only ever need this.  Launch it
  with `run_seqNdisplayR_app()`.
- **R API / scripting (advanced).**  The same underlying
  `seqNdisplay()` function is exported for pipelines and batch plotting.
  Useful when you want to generate many plots from a script, or embed
  plotting in a larger workflow.

Both paths share the same Excel template and produce identical plots
from identical inputs.  This vignette walks the Shiny path first, then
shows the scripting equivalent so you can see how the Excel columns
map to R structures.


## Installation

Full install instructions with troubleshooting live in the
[README](https://github.com/slaish/seqNdisplayR#installation).  The
recipe below is the shortest happy path.

### 1. GLPK (system dependency)

The annotation-name placement uses the GNU Linear Programming Kit via
`ROI.plugin.glpk` / `ompr.roi`.  Install once, at the OS level:

| OS                    | Command                                              |
|-----------------------|------------------------------------------------------|
| Debian / Ubuntu       | `sudo apt-get install libglpk-dev`                   |
| macOS (Homebrew)      | `brew install glpk`                                  |
| Windows               | bundled with the Rtools toolchain (R 4.0+)           |

### 2. Bioconductor dependencies

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(c("rtracklayer", "GenomicRanges", "BiocGenerics",
                       "GenomeInfoDb", "S4Vectors", "IRanges", "limma"),
                     update = FALSE, ask = FALSE)
```

### 3. The package itself

```{r, eval=FALSE}
if (!requireNamespace("remotes", quietly = TRUE))
    install.packages("remotes")
remotes::install_github("slaish/seqNdisplayR", ref = "main")
```

This pulls in the GitHub-only `bwimport` automatically via the
package's `Remotes:` field.  On Windows the
[`pak`-based recipe](https://github.com/slaish/seqNdisplayR#windows-friendly-alternative-pak--binary-only-install)
is cleaner and avoids source-build stalls (`survival` etc.); see the
README.


## Quick start (Shiny)

Launch the app:

```{r, eval=FALSE}
library(seqNdisplayR)
run_seqNdisplayR_app()
```

A browser tab opens with the app.  The layout has a left sidebar (data
loading + controls) and a right main pane (the plot).

- **Load Excel** — click `Browse...` and pick a sample sheet.  For a
  first look, use the shipped simple example:
  `system.file("extdata", "sNdR_sample_example_simple.xlsx",
   package = "seqNdisplayR")`.  A more elaborate example lives at
  `sNdR_sample_example_elaborate.xlsx`.
- **Data Overview** — a tree of datasets / subgroups / samples with
  checkboxes.  Uncheck rows to hide them from the plot.
- **Plot Segments & Strands** — a drag-reorderable table controlling
  the top-to-bottom order of the plot (header, scale, annotations,
  spacers, individual datasets).  Add / remove rows with the buttons
  on each row.  The strand-display controls (`bothstrands`,
  `intermingled`, the intermingled colour picker, `neg_as_neg`,
  `reverse_strand`) live here too.
- **Track Colors** — per-sample colour pickers, populated from the
  Excel or the last edit.
- **Plot Layout** — panel widths, margins, scale-panel behaviour,
  "Use full panel width", and the auto-widening toggle for label-heavy
  panels.
- **IGV Export** — writes an IGV session XML for the currently loaded
  configuration; you'll be prompted for a target genome (default `hg38`).
- **Console (click to show / hide)** — a collapsible bottom panel with
  package-side messages.  Useful when debugging a plot that doesn't
  render.

To draw a plot: type a gene symbol (e.g. `LMO4`) or coordinates
(`chr1:87297784-87379607`) into **Locus of Interest**, then click
**Draw Plot**.  On first plot the tracks are fetched from their remote
URLs; subsequent plots at the same or different loci reuse the
per-session bigwig cache and are much faster.


## The Excel template as the bridge

An Excel sample sheet has four sheets:

- **SAMPLES** — one row per bigwig track: `color`, `bigwig_directory`,
  `bigwig_file`, `strand`, `batch`, `dataset`, `subgroup_1` (and
  optionally `subgroup_2`, `subgroup_3`).  The `dataset` /
  `subgroup_*` columns define the nesting the plot uses.
- **DATASET_OPTIONS** — per-dataset parameters: `bin_stats`,
  `log2transform`, `pseudoCount`, `batchCorrect`, `enhance_signals`,
  etc.
- **ANNOTATIONS** — one row per BED file with layout options
  (`incl_feature_names`, `annotation_packing`, `annot_cols`, ...).
- **GLOBAL_OPTIONS** — layout / plot-level options that would otherwise
  be arguments to `seqNdisplay()` (margins, `neg_as_neg`,
  `bothstrands`, `intermingled`, etc.).

The Shiny app lets you round-trip: load an xlsx, edit interactively,
export the current state back to xlsx with **Save Excel** (calls
`Session2xlsx()` under the hood).  The exported xlsx is the
authoritative record of a plot's configuration.

You can also inspect and convert between formats programmatically:

```{r, eval=FALSE}
xl_path <- system.file("extdata", "sNdR_sample_example_simple.xlsx",
                       package = "seqNdisplayR")
session <- LoadExcel(xl_path)      # -> seqNdisplayRSession
print(session)                     # summary
Session2Df(session)                # tabular view (data.frame)
Session2xlsx(session, "my_edited.xlsx")
```


## R API / scripting

Everything the Shiny app does is available as R functions.  Below is
the CLI equivalent of loading the shipped simple example and plotting
`LMO4`.

### Sample structure

A nested list mirrors the SAMPLES sheet: dataset → subgroup → samples.
For the shipped simple example (three stranded datasets with siCTRL /
siEXOSC3 subgroups, plus an unstranded ChIP dataset):

```{r}
samples <- list(
  "3-seq"   = list('siCTRL' = 'siGFP', 'siEXOSC3' = 'siRRP40'),
  "TT-seq"  = list('siCTRL' = 'siGFP', 'siEXOSC3' = 'siRRP40'),
  "RNA-seq" = list('siCTRL' = 'siGFP', 'siEXOSC3' = 'siRRP40'),
  "ChIP-seq" = c('ChIP')
)
```

### Colors

Same nesting; deepest leaf is a named vector.

```{r}
colors <- list(
  "3-seq"   = list('siCTRL' = c('siGFP' = '#505160'),
                   'siEXOSC3' = c('siRRP40' = '#68829E')),
  "TT-seq"  = list('siCTRL' = c('siGFP' = '#505160'),
                   'siEXOSC3' = c('siRRP40' = '#68829E')),
  "RNA-seq" = list('siCTRL' = c('siGFP' = '#505160'),
                   'siEXOSC3' = c('siRRP40' = '#68829E')),
  "ChIP-seq" = c('ChIP' = '#B0B0B0')
)
```

### BigWig directories and files

Directories per dataset (public URLs; use `https://user:password@host/...`
for password-protected lab servers):

```{r}
bigwig_dirs <- c(
  "3-seq"    = 'http://genome-ftp.mbg.au.dk/public/THJ/seqNdisplayR/examples/tracks/HeLa_3pseq/',
  "TT-seq"   = 'http://genome-ftp.mbg.au.dk/public/THJ/seqNdisplayR/examples/tracks/HeLa_TTseq/',
  "RNA-seq"  = 'http://genome-ftp.mbg.au.dk/public/THJ/seqNdisplayR/examples/tracks/HeLa_RNAseq/',
  "ChIP-seq" = 'http://genome-ftp.mbg.au.dk/public/THJ/seqNdisplayR/examples/tracks/HeLa_ChIPseq/'
)
```

Individual track filenames, split by strand (omit the `-` entry for
unstranded data; the ChIP-seq dataset is unstranded):

```{r}
bigwigs <- list(
  '+' = list(
    "3-seq"   = list('siCTRL'   = list('siGFP'  = c('siGFP_xPAP_in_batch1_plus.bw',
                                                    'siGFP_xPAP_in_batch2_plus.bw',
                                                    'siGFP_xPAP_in_batch3_plus.bw')),
                     'siEXOSC3' = list('siRRP40'= c('siRRP40_xPAP_in_batch1_plus.bw',
                                                    'siGFP_xPAP_in_batch2_plus.bw',
                                                    'siRRP40_xPAP_in_batch3_plus.bw'))),
    "TT-seq"  = list('siCTRL'   = list('siGFP'  = c('L_EGFP_rep1_tt_corr_ff_noJncReads_plus.bw',
                                                    'L_EGFP_rep2_tt_corr_ff_noJncReads_plus.bw')),
                     'siEXOSC3' = list('siRRP40'= c('L_RRP40_rep1_tt_corr_ff_noJncReads_plus.bw',
                                                    'L_RRP40_rep2_tt_corr_ff_noJncReads_plus.bw'))),
    "RNA-seq" = list('siCTRL'   = list('siGFP'  = c('T_EGFP_rep1_tt_corr_plus.bw',
                                                    'T_EGFP_rep2_tt_corr_plus.bw')),
                     'siEXOSC3' = list('siRRP40'= c('T_RRP40_rep1_tt_corr_plus.bw',
                                                    'T_RRP40_rep2_tt_corr_plus.bw'))),
    "ChIP-seq" = list('ChIP' =
        c('GSM2642506_WIGfs_Hela-H9_WT_siFFL_Pol_II_N20_MA733_bin50_Scaled_BGSub_Hg38.bw',
          'GSM2642508_WIGfs_Hela-H9_WT_siFFL_Pol_II_N20_MA736_bin50_Scaled_BGSub_Hg38.bw'))
  ),
  '-' = list(
    "3-seq"   = list('siCTRL'   = list('siGFP'  = c('siGFP_xPAP_in_batch1_minus.bw',
                                                    'siGFP_xPAP_in_batch2_minus.bw',
                                                    'siGFP_xPAP_in_batch3_minus.bw')),
                     'siEXOSC3' = list('siRRP40'= c('siRRP40_xPAP_in_batch1_minus.bw',
                                                    'siGFP_xPAP_in_batch2_minus.bw',
                                                    'siRRP40_xPAP_in_batch3_minus.bw'))),
    "TT-seq"  = list('siCTRL'   = list('siGFP'  = c('L_EGFP_rep1_tt_corr_ff_noJncReads_minus.bw',
                                                    'L_EGFP_rep2_tt_corr_ff_noJncReads_minus.bw')),
                     'siEXOSC3' = list('siRRP40'= c('L_RRP40_rep1_tt_corr_ff_noJncReads_minus.bw',
                                                    'L_RRP40_rep2_tt_corr_ff_noJncReads_minus.bw'))),
    "RNA-seq" = list('siCTRL'   = list('siGFP'  = c('T_EGFP_rep1_tt_corr_minus.bw',
                                                    'T_EGFP_rep2_tt_corr_minus.bw')),
                     'siEXOSC3' = list('siRRP40'= c('T_RRP40_rep1_tt_corr_minus.bw',
                                                    'T_RRP40_rep2_tt_corr_minus.bw')))
  )
)
```

### Parameters

Per-dataset transformation and replicate-handling settings.  Any field
omitted here falls back to `DefaultParameters()`.

```{r}
parameters <- list(
  "3-seq"    = list(bin_stats = 'mean', log2transform = FALSE, pseudoCount = 1,
                    batchCorrect = TRUE, batch = c(1, 2, 3, 1, 2, 3),
                    calcMean = TRUE, enhance_signals = TRUE, negValsSet0 = TRUE),
  "TT-seq"   = list(bin_stats = 'mean', log2transform = FALSE, pseudoCount = 1,
                    batchCorrect = TRUE, batch = c(1, 2, 1, 2),
                    calcMean = TRUE, enhance_signals = FALSE, negValsSet0 = TRUE),
  "RNA-seq"  = list(bin_stats = 'mean', log2transform = FALSE, pseudoCount = 1,
                    batchCorrect = TRUE, batch = c(1, 2, 1, 2),
                    calcMean = TRUE, enhance_signals = FALSE, negValsSet0 = TRUE),
  "ChIP-seq" = list(bin_stats = 'mean', log2transform = FALSE, pseudoCount = 1,
                    batchCorrect = FALSE, calcMean = TRUE,
                    enhance_signals = FALSE, negValsSet0 = TRUE)
)
```

Field notes: `bin_stats` picks the per-bin aggregation (`mean`, `sum`,
etc.); `enhance_signals` applies a subtle non-linear boost so low-signal
regions become visible; `batchCorrect` runs `limma::removeBatchEffect`
across replicates when `batch` groups them; `calcMean = FALSE` displays
individual replicates rather than the per-condition mean.

### Annotations

Pre-load once with `ReadInAnnotations()` so the same GRanges are reused
across all plots (avoids downloading the BED files on every call):

```{r}
annotations <- ReadInAnnotations(list(
  'gencode v38' = 'http://genome-ftp.mbg.au.dk/public/THJ/seqNdisplayR/examples/annotations/gencode.v38.annotation.bed',
  'in-house'    = 'http://genome-ftp.mbg.au.dk/public/THJ/seqNdisplayR/examples/annotations/HeLa_major_isoform_hg38_gc34.bed',
  'ChIP-peaks'  = 'http://genome-ftp.mbg.au.dk/public/THJ/seqNdisplayR/examples/annotations/RNAPII_ChIP_peaks.bed'
))
```

### Plot

Direct call:

```{r}
seqNdisplay(samples, colors, bigwig_dirs, bigwigs, parameters,
            annots = annotations, feature = 'LMO4')
```

Or via a `seqNdisplayRSession` object, which bundles the arguments and
supports `plot()` / `print()` / xlsx round-trip:

```{r}
session <- seqNdisplayRSession(samples     = samples,
                               colors      = colors,
                               bigwig_dirs = bigwig_dirs,
                               bigwigs     = bigwigs,
                               parameters  = parameters,
                               annotations = annotations)
print(session)                          # summary
plot(session, feature = 'LMO4')         # same as seqNdisplay(...)
plot(session, feature = 'chr1:87297784-87379607')  # coordinates work too
```

Both routes give the same plot as clicking **Draw Plot** in the Shiny
app with the same Excel loaded.


## IGV import / export (v2.0.0)

`seqNdisplayR` reads and writes IGV session XML files so you can jump
between the two tools without re-authoring configurations.

**Export.**  From an existing `seqNdisplayRSession`:

```{r}
Session2IGV(session,
            genome = 'hg38',
            out    = 'my_session.xml')
```

The Shiny app's IGV Export button wraps the same call and prompts for
the target genome via a small modal (common identifiers listed;
defaults to `hg38`).

**Import.**  From an IGV session:

```{r}
imported <- IGV2Session('my_session.xml')
print(imported)
```

When `IGV2Session()` can't infer a required field from the XML
(synthetic `solo_N` / `group_N` dataset names, missing batches or
subgroups), it writes `<FILL_ME: ...>` sentinel strings into the
generated xlsx.  Loading that xlsx in the Shiny app auto-runs
`CheckSampleFile()`, which surfaces the sentinels as warnings pointing
at each row that needs completion.

For a walk-through of a full XLSX ⇄ IGV round-trip see
`inst/docs/igv_import.md` in the source.


## Performance tips

- **Parallel bigwig fetch.**  Install `future` and `future.apply` to
  enable parallel per-track loading.  seqNdisplayR detects them and
  fans out fetches across a persistent worker pool (cross-platform,
  Windows-compatible).  Falls back to sequential otherwise.
  Turn off explicitly with
  `options(seqNdisplayR.disable_parallel_bw = TRUE)`.
- **Session-level result caches.**  The bigwig fetcher caches results
  by `(file, chrom, start, end)`; replotting the same locus, or
  overlapping loci for the same bigwigs, is near-instant.  Free memory
  with `clear_bigwig_cache()`, inspect with `bigwig_cache_size()`.
  The annotation-processing pipeline has a similar cache
  (`clear_annotation_cache()` / `annotation_cache_size()`).
- **Fast Windows remote fetch.**  Windows plots require
  `bwimport (>= 0.2.3)` for the byte-range fast path;  earlier
  versions mangled URLs and downloaded every bigwig in full.  The
  pin in `DESCRIPTION` enforces this, but check with
  `packageVersion("bwimport")` if you inherited an old install.
- **libBigWig quiet mode.**  Transient first-attempt failures
  (`[bwHdrRead]` / `[bwOpen]` chatter) are suppressed by default
  (seqNdisplayR sets `BWIMPORT_QUIET = "1"` in `.onLoad`).  Set
  `Sys.setenv(BWIMPORT_QUIET = "")` before `library(seqNdisplayR)` to
  see the original diagnostics.


## Troubleshooting

The full list is in the
[README](https://github.com/slaish/seqNdisplayR#troubleshooting).  Top
hits:

- **Install fails at lazy-load with
  `class "Seqinfo" is not exported by namespace:GenomeInfoDb`.**
  Bioconductor version mismatch.  Sync everything to the current
  release: `BiocManager::install(ask = FALSE, update = TRUE)`, then
  retry.
- **`"GLPK not found"` / `ROI.plugin.glpk` install errors.**
  Install GLPK at the OS level (§Installation §1), then reinstall
  `ROI.plugin.glpk`.
- **Shiny app fails to launch with
  `"there is no package called 'XYZ'"`.**  Reinstall with
  `remotes::install_github("slaish/seqNdisplayR", dependencies = TRUE,
   upgrade = "never")`.
- **`install_github()` was deprecated in devtools 2.5.0** — harmless
  notice; the underlying call still works.  Prefer
  `pak::pak("slaish/seqNdisplayR@main")` on Windows.
- **Windows plot is very slow.**  Check `packageVersion("bwimport") >=
  "0.2.3"` and consider installing `future` + `future.apply` for
  parallel fetch.


## Where to go next

- The [README](https://github.com/slaish/seqNdisplayR/blob/main/README.md)
  has the full install / troubleshooting reference.
- [`docs/deployment.md`](https://github.com/slaish/seqNdisplayR/blob/main/docs/deployment.md)
  covers hosting the Shiny app for collaborators (Shiny Server,
  Docker + ShinyProxy).
- [`inst/docs/igv_import.md`](https://github.com/slaish/seqNdisplayR/blob/main/inst/docs/igv_import.md)
  walks through the IGV round-trip in detail.
- [`NEWS.md`](https://github.com/slaish/seqNdisplayR/blob/main/NEWS.md)
  documents every change in v2.0.0 with the underlying rationale.
